---
title: "Gerrymandering Project Dashboard"
format: dashboard
---


```{r}
#Set up 
library(tidyverse)
library(ggplot2)
library(sf)
library(janitor)

ab604_results <- read_csv("data/ab604_cd_results.csv")
results_2024 <- read_csv("data/g24_sov_by_g24_svprec_clean.csv")

ab604_results |> 
    clean_names()
results_2024 |> 
    clean_names()

results_2024 <- results_2024 |>
    rename(dem_total = demvote, rep_total = repvote)

ab604_results <- ab604_results |>
    rename(dem_total = uspdem01, rep_total = usprep01)


```
```{r}
#Metric Functions
wasted_votes <- function(votes_a, votes_b) {
total <- votes_a + votes_b
needed <- floor(total / 2) + 1
ifelse(votes_a > votes_b, votes_a - needed, votes_a)
}

compute_metrics <- function(df, plan_name) {
    df2 <- df |>
    mutate( total = dem_total + rep_total, dem_share = dem_total / total) |>
    filter(total > 0)

n_seats <- nrow(df2)
dem_seats <- sum(df2$dem_total > df2$rep_total)
rep_seats <- sum(df2$rep_total > df2$dem_total)

dem_seat_share <- dem_seats / n_seats

mean_dem_share <- mean(df2$dem_share)
median_dem_share <- median(df2$dem_share)
mean_median <- mean_dem_share - median_dem_share

wasted_dem <- wasted_votes(df2$dem_total, df2$rep_total)
wasted_rep <- wasted_votes(df2$rep_total, df2$dem_total)

total_wasted_dem <- sum(wasted_dem)
total_wasted_rep <- sum(wasted_rep)
total_two_party_votes <- sum(df2$total)

efficiency_gap <- (total_wasted_rep - total_wasted_dem) / total_two_party_votes

tibble(plan = plan_name, n_seats = n_seats, dem_seats = dem_seats, rep_seats = rep_seats, dem_seat_share = dem_seat_share, mean_median = mean_median, efficiency_gap = efficiency_gap)
}

metrics_2024 <- compute_metrics(results_2024, "Current 2024 map")
metrics_ab604 <- compute_metrics(ab604_results, "AB 604 map")

metrics_all <- bind_rows(metrics_2024, metrics_ab604)
```
```{r}
#Read Shape Files 
cd2024_shp <- st_read("data/shapefiles/tl_2024_06_cd119/tl_2024_06_cd119.shp")

cd2024_shp |>
    clean_names() |>
    st_transform(3310)

cd2024_shp <- cd2024_shp |>
    mutate(GEOID = as.numeric(GEOID))

cd2024_map <- cd2024_shp |>
    left_join(results_2024, by = c("GEOID" = "cddist")) |> 
    mutate(total_votes = dem_total + rep_total, dem_share = dem_total / total_votes, winner = case_when(is.na(dem_share) ~ "No Data", dem_share > 0.5 ~ "Democrat", dem_share < 0.5 ~ "Republican", TRUE ~ "Tie"))

ab604_shp <- st_read("data/shapefiles/AB604/AB604.shp")

ab604_shp |> 
    clean_names() |>
    st_transform(3310)

ab604_map <- ab604_shp |>
    left_join(ab604_results, by = c("DISTRICT" = "district")) |> 
    mutate(total_votes = dem_total + rep_total, dem_share = dem_total / total_votes, winner = case_when(is.na(dem_share) ~ "No Data", dem_share > 0.5 ~ "Democrat", dem_share < 0.5 ~ "Republican", TRUE ~ "Tie"))
```
```{r}
#Maps 
#Current 2024 Districts 
cd2024_map |> 
    ggplot() +
    geom_sf(aes(fill = winner), color = NA) +
    scale_fill_manual( values = c("Democrat" = "#3182bd", "Republican" = "#de2d26", "Tie" = "grey60", "No Data" = "grey85")) +
    labs(title = "2024 Results: Current Map", fill = "Winner") +
    theme_minimal()
#AB604 Districts
ab604_map |> 
    ggplot() +
    geom_sf(aes(fill = winner), color = NA) +
    scale_fill_manual(values = c("Democrat" = "#3182bd","Republican" = "#de2d26","Tie" = "grey60", "No Data" = "grey85")) +
    labs(title = "Hypothetical 2024 Results – AB 604", fill = "Winner") +
    theme_minimal()
```

Methodology
For this project I used the SR precinct data on voting from the Statewide Database for the 2024 US House election in California. I joined the SR precinct votes to SR precinct geometries using the sf library. I compared the status quo map to the AB 604 map. For the current map I aggregated percent votes to 2024 congressional districts. For the AB 604 map I used area-weighted interpolation to estimate district level votes. I then computes the mean median vote share and the efficiency gap. 